Theory about Normal Basis on $F(2^k)$

### 2016-12-31 dissertation/submission/appendix1.tex
dissertation milestone: submitting to thesis office

Ultimately, since there always exist at least one element fulfilling requirements in Theorem \ref{thm:1}
or Theorem \ref{thm:2}, we obtain following theorem:
\begin{Theorem}[Normal Basis Theorem over Finite Fields]
There always exists a normal basis of $\F_{p^n}$ over $\F_p$.
\end{Theorem}
\section{Construction of General Normal Bases}
After proving the existence of normal bases, the upcoming question is that how to find such a normal basis/element.
For general normal basis identification, there are two methods widely used: the L\"uneburg's algorithm
and Lenstra's algorithm.\\
\textbf{L\"uneburg's Algorithm:}
\begin{enumerate}[{1)}]
\item Randomly pick an element $\alpha$ from $\F_{p^n}$. For each $i = 0,1,\dots,n-1$, 
compute $\sigma$-Order $f_i = Ord_{\alpha^i}(x)$. Then $x^n - 1 = lcm(f_0,f_1,...,f_{n-1})$.
\item Apply factor refinement to set $\{f_0,\dots,f_{n-1}\}$ and obtain $f_i = \prod_{1\leq j\leq r} g_{j}^{e_{ij}}, i = 0,1,\dots,n-1$.
We can write the result as an $i\times j$ matrix.
\item For each $j$, find an index $i_j$ (denote as $i(j)$) such that $e_{ij}$ is maximum in the $j$-th column.
\item Let $h_j = f_{i(j)}/g_{j}^{e_{i(j)j}}$, take $\beta_j = h_j(\sigma)\alpha^{i(j)}$. Then
$$\beta = \displaystyle\sum_{j=1}^{r} \beta_j$$
is a normal element.
\end{enumerate}

@@

L\"uneburg's algorithm is an analytic method, which requires relatively high computational complexity 
mainly because of the factorization in its second step. To overcome the high cost, an inductive method 
is proposed, which is the Lenstra's algorithm. It allows for setup of heuristics to accelerate the procedure.
Before introducing the details of the algorithm, we demonstrate two preliminary lemmas.
\begin{Lemma}
\label{lem:4}
For an arbitrary element $\theta \in \F_{p^n}$ that $Ord_\theta(x) \neq x^n - 1$,
let $g(x) = (x^n - 1)/Ord_\theta(x)$. There exists another element $\beta$ such that $g(\sigma)\beta = \theta$.
\end{Lemma}
\begin{Proof}
Assume $\gamma$ is the desired normal element. From the definition of normal element we derive 
$$\exists f(x)\in \F_{p^n}[x],~~f(\sigma)\gamma = \theta$$

@@

Based on Lemmas \ref{lem:4} and \ref{lem:5}, the Lenstra's algorithm is described below:\\
\textbf{Lenstra's Algorithm}
\begin{enumerate}[{1)}]
\item Take an arbitrary element $\theta \in \F_{p^n}$, determine $Ord_\theta(x)$.
\item If $Ord_\theta(x) = x^n - 1$ then algorithm terminates and return $\theta$ as a normal element.
\item Otherwise, compute $g(x) = (x^n - 1)/Ord_\theta(x)$, and solve $\beta$ from $g(\sigma)\beta = \theta$.
\item Determine $Ord_\beta(x)$. If $deg(Ord_\beta(x)) > deg(Ord_\theta(x))$ then replace $\theta$ by $\beta$ and go to 2);
otherwise if $deg(Ord_\beta(x)) \leq deg(Ord_\theta(x))$ then find a non-zero element $\eta$ such that $g(\sigma)\eta = 0$,
replace $\theta$ by $\theta + \eta$ and determine the order of new $\theta$, and go to 2).
\end{enumerate}

>>>>

Ultimately, since there always exist at least one element fulfilling requirements in Theorem \ref{thm:1}
or Theorem \ref{thm:2}, we obtain following theorem:
\begin{Theorem}[Normal Basis Theorem over Finite Fields]
There always exists a normal basis of $\F_{p^n}$ over $\F_p$.
\end{Theorem}
\section{Construction of General Normal Bases}
After proving the existence of normal bases, the upcoming question is that how to find such a normal basis/element.
For general normal basis identification, there are two methods widely used: the L\"uneburg's algorithm
and Lenstra's algorithm.

\subsection{L\"uneburg's Algorithm}
L\"uneburg's algorithm can be described with following steps:

\begin{enumerate}[{1)}]
\item Randomly pick an element $\alpha$ from $\F_{p^n}$. For each $i = 0,1,\dots,n-1$, 
compute $\sigma$-Order $f_i = Ord_{\alpha^i}(x)$. Then $x^n - 1 = lcm(f_0,f_1,...,f_{n-1})$.
\item Apply factor refinement to set $\{f_0,\dots,f_{n-1}\}$ and obtain $f_i = \prod_{1\leq j\leq r} g_{j}^{e_{ij}}, i = 0,1,\dots,n-1$.
We can write the result as an $i\times j$ matrix.
\item For each $j$, find an index $i_j$ (denote as $i(j)$) such that $e_{ij}$ is maximum in the $j$-th column.
\item Let $h_j = f_{i(j)}/g_{j}^{e_{i(j)j}}$, take $\beta_j = h_j(\sigma)\alpha^{i(j)}$. Then
$$\beta = \displaystyle\sum_{j=1}^{r} \beta_j$$
is a normal element.
\end{enumerate}

@@

\subsection{Lenstra's Algorithm}
L\"uneburg's algorithm is an analytic method, which requires relatively high computational complexity 
mainly because of the factorization in its second step. To overcome the high cost, an inductive method 
is proposed, which is the Lenstra's algorithm. It allows for setup of heuristics to accelerate the procedure.
Before introducing the details of the algorithm, we demonstrate two preliminary lemmas.
\begin{Lemma}
\label{lem:4}
For an arbitrary element $\theta \in \F_{p^n}$ that $Ord_\theta(x) \neq x^n - 1$,
let $g(x) = (x^n - 1)/Ord_\theta(x)$. There exists another element $\beta$ such that $g(\sigma)\beta = \theta$.
\end{Lemma}
\begin{Proof}
Assume $\gamma$ is the desired normal element. From the definition of normal element we derive 
$$\exists f(x)\in \F_{p^n}[x],~~f(\sigma)\gamma = \theta$$

@@

Based on Lemmas \ref{lem:4} and \ref{lem:5}, the Lenstra's algorithm is described below:

\begin{enumerate}[{1)}]
\item Take an arbitrary element $\theta \in \F_{p^n}$, determine $Ord_\theta(x)$.
\item If $Ord_\theta(x) = x^n - 1$ then algorithm terminates and return $\theta$ as a normal element.
\item Otherwise, compute $g(x) = (x^n - 1)/Ord_\theta(x)$, and solve $\beta$ from $g(\sigma)\beta = \theta$.
\item Determine $Ord_\beta(x)$. If $deg(Ord_\beta(x)) > deg(Ord_\theta(x))$ then replace $\theta$ by $\beta$ and go to 2);
otherwise if $deg(Ord_\beta(x)) \leq deg(Ord_\theta(x))$ then find a non-zero element $\eta$ such that $g(\sigma)\eta = 0$,
replace $\theta$ by $\theta + \eta$ and determine the order of new $\theta$, and go to 2).
\end{enumerate}

### 2016-12-31 dissertation/submission/conclude.tex
dissertation milestone: submitting to thesis office

\section{Future Work}
In this section, we highlight some research problems that deserve further study.
\subsection{Multivariate Polynomial Ideal based FSM Traversal}
In Chapter \ref{ch:reacha}, we always use a single word-level variable $T$ to denote the
next state. However, in some situations we need to keep recording the relations between $T$ and
inputs, {\it i.e.} the reached states contain multivariate polynomials in $T,S$ and PI $x$. 
While the elimination and set union/intersection are compatible with multiple NS variables,
the set complement requires an extension of Theorem \ref{thm:quotient} on ideals with 
multivariate polynomial generators. We conjecture as follows:

@@

The following example illustrates a different operation from Example \ref{ex:SMPO}.
\begin{Example}
In Figure \ref{fig:fsm}
$\{s_0, s_1\}$ are state/pseudo inputs, $\{t_0,t_1\}$ are state/pseudo outputs, and there is a primary input (1-bit) 
$x$. We propose a new algorithm by modifying Algorithm \ref{alg:univa}:

@@

The inputs of this algorithm includes transition polynomial (result of word-level abstraction) and initial 
states description ideal, which contains 2 generators corresponds to constraints of PI and combinational input $S$. 
For example, $\langle S+1+\alpha, x^2+x\rangle$ means initial state$=\{11\}$,
$\langle S+x\cdot\alpha, x^2+x\rangle$ means initial states$=\{00,10\}$).

Transition polynomial calculation uses the abstraction and bit-to-level conversion method from 
Chapter \ref{ch:reacha}. After Constructing
an elimination ideal, we impose RATO such that \emph{reverse\ topo\ order\ ckt\ variables }$> T > S > x$, the reduction
remainder has the form $T+\mathcal F(s_0,s_1,x)$. From word definition $S+s_0+s_1\alpha$ we get
$$s_0 = \alpha S^2+ (1+\alpha)S, s_1 = S^2+S$$
Do substitution, the transition polynomial of example ckt is 
$$f_T = T+S^3\cdot x+\alpha S^3+(1+\alpha)S^2\cdot x+S^2+S\cdot x+(1+\alpha)x+1$$

Assume we start from state $\{11\}$. In the first iteration, the reached state is $\{01\}$. Line 4 is to compose an
ideal with 2 generators from $from^0$ and transition polynomial $f_T$, compute its Gr\"obner basis. Note this ideal
has the form
\begin{equation}
I_{tran} = \left\{
             \begin{array}{c}
             T+\mathcal F(S,x) \\
             S + \mathcal G(x) \\
             v_x
             \end{array}  
        \right.
\end {equation}
$v_x$ is a polynomial containing only $q$-bit PI $x$, initially it should be vanishing polynomial $x^{q_1}-x$, 
with the program executing it may be factorized.

Consider Buchberger's algorithm, all generators' leading terms are relatively prime, so it is a GB itself. 
Then we need to reduce this GB, we will find out $S + \mathcal G(x)$ could (possibly) be reduced by $v_x$,
and $T+\mathcal F(S,x)$ will definitely reduced by $S + \mathcal G(x)$. So, at last we will get a polynomial
$T + \mathcal F'(x)$ in reduce GB. We can include this polynomial and $v_x$, exclude the polynomial containing
$S$ ({\it i.e.} $\mathcal H(S)$ in algorithm), to compose an ideal representing \emph{next states} $to^i$. 
In iteration 1, result is $to^1 = \langle T+1, x^2+x\rangle$

@@

Line 6 is ideals' sum (intersection of their varieties), it is done by combining all generators
from 2 ideals and compute GB. For first iteration result is $GB(\langle T+1,T^3+(1+\alpha)T^2+\alpha T, x^2+x\rangle) = \langle T+1, x^2+x\rangle$ representing $\{01\}$.

Line 7 is ideals' product (union of their varieties), it is done by multiplying all pairs of
generators from both ideal. For first iteration result is $GB(\langle (T+1)(T+1+\alpha),
(T+1)(x^2+x), (T+1+\alpha)(x^2+x), (x^4+x)\rangle) = \langle T^2+\alpha T+(1+\alpha), x^2+x\rangle$ representing $\{01,11\}$.

The traversal will run 3 iterations and terminate at 4th iteration. I list all intermediate 
results below:
\begin{itemize}
\item Iteration 1: $from^0 = \langle S+1+\alpha, x^2+x\rangle, to^1 = \langle T+1, x^2+x\rangle,
 reached = \langle T^2+\alpha T+(1+\alpha), x^2+x\rangle$
\item Iteration 2: $from^1 = \langle S+1, x^2+x\rangle, to^2= \langle T+\alpha, x^2+x\rangle,
reached = \langle T^3+1, x^2+x\rangle$
\item Iteration 3: $from^2 = \langle S+\alpha, x^2+x\rangle, to^3 = \langle T+\alpha x, x^2+x
\rangle, reached = \langle T^3\cdot x+x, T^4+T, x^2+x\rangle$
\item Iteration 4: $from^3 = \langle S, x\rangle, to^4 = \langle T+1, x\rangle, new = \langle1\rangle$
\end{itemize}
The final reachable states are represented by multivariate polynomial ideal $\langle T^3\cdot x+x, T^4+T, x^2+x\rangle$,
which denotes $\{00,01,10,11\}$.
\end{Example}

@@

Figure \ref{fig:ZDD} shows a ZDD representing 
Equation \ref{eqn:chainOR} with size $2\times4 -1 = 7$, which is linear in the number of variables.
The reduction process using ZDDs can be executed as in \cite{polybori:2009}. The graphical illustration 
of the remainder in ZDDs is also shown in Figure \ref{fig:ZDD}.

>>>>

\section{Future Work}
In this section, we highlight some research problems that deserve further study.
\subsection{Multivariate Polynomial Ideal based FSM Traversal}
In Chapter \ref{ch:reacha}, we always use a single word-level variable $T$ to denote the
next state. However, in some situations we need to keep recording the relations between $T$ and
the inputs, {\it e.g.} the reached states contain multivariate polynomials in $T$, as well as 
PS variable $S$ and PI $x$. 
While the elimination and set union/intersection are compatible with multivariate polynomials,
the set complement requires an extension of Theorem \ref{thm:quotient} on ideals with 
multivariate polynomial generators. We conjecture as follows:

@@

The following example illustrates a different operation from Example \ref{ex:SMPO}.
\begin{Example}
In Figure \ref{fig:fsm}, we denote that
$\{s_0, s_1\}$ are state/pseudo inputs, $\{t_0,t_1\}$ are state/pseudo outputs, and there is a primary input (1-bit) 
$x$. We propose a new algorithm (Algorithm \ref{alg:multi}) by modifying Algorithm \ref{alg:univa}.

@@

The inputs of this algorithm includes the transition polynomial (result of word-level abstraction) and initial 
states description ideal, which contains 2 generators corresponding to constraints of PI and combinational input $S$. 
For example, $\langle S+1+\alpha, x^2+x\rangle$ means initial state$=\{11\}$,
$\langle S+x\cdot\alpha, x^2+x\rangle$ means initial states$=\{00,10\}$).

Transition polynomial calculation uses the abstraction and bit-to-word conversion method from 
Chapter \ref{ch:reacha}. After Constructing
an elimination ideal, we impose RATO such that \textit{reverse\ topo\ order\ ckt\ variables }$> T > S > x$, the reduction
remainder has the form $T+\mathcal F(s_0,s_1,x)$. According to the bit-to-word correspondence $S+s_0+s_1\alpha$ we get
$$s_0 = \alpha S^2+ (1+\alpha)S, s_1 = S^2+S$$
After substitution, we obtain the transition polynomial of example circuit: 
$$f_T = T+S^3\cdot x+\alpha S^3+(1+\alpha)S^2\cdot x+S^2+S\cdot x+(1+\alpha)x+1$$

Assume the initial state is $\{11\}$. In the first iteration, the reached state is $\{01\}$. Line 4 is to compose an
ideal with 2 generators from $from^0$ and transition polynomial $f_t$, compute its Gr\"obner basis. Note this ideal
has the form
\begin{equation}
I_{tran} = \left\{
             \begin{array}{c}
             T+\mathcal F(S,x) \\
             S + \mathcal G(x) \\
             v_x
             \end{array}  
        \right.
\end {equation}
$v_x$ is a polynomial containing only $q_1$-bit PI $x$, initially it should be vanishing polynomial $x^{q_1}-x$, 
with the program executing it may be factorized.

Considering the Buchberger's algorithm, all generators' leading terms are relatively prime, so $I_{tran}$ is a GB itself. 
Furthermore, generator $S + \mathcal G(x)$ could (possibly) be reduced by $v_x$,
and $T+\mathcal F(S,x)$ will definitely reduced by $S + \mathcal G(x)$. In the end we obtain a polynomial
$T + \mathcal F'(x)$ in the reduced GB. We can then include this polynomial and $v_x$ and exclude the polynomial containing
$S$ ({\it i.e.} $\mathcal H(S)$ in Algorithm \ref{alg:multi}), to compose an ideal representing \emph{next states} $to^i$. 
In iteration 1, the result is $to^1 = \langle T+1, x^2+x\rangle$

@@

Line 6 is ideals' sum (intersection of their varieties), it is done by combining all generators
from 2 ideals and compute GB. In the first iteration, the result is $GB(\langle T+1,T^3+(1+\alpha)T^2+\alpha T, x^2+x\rangle) = \langle T+1, x^2+x\rangle$ representing $\{01\}$.

Line 7 is ideals' product (union of their varieties), it is done by multiplying all pairs of
generators from both ideal. For the first iteration it gives the result as $GB(\langle (T+1)(T+1+\alpha),
(T+1)(x^2+x), (T+1+\alpha)(x^2+x), (x^4+x)\rangle) = \langle T^2+\alpha T+(1+\alpha), x^2+x\rangle$ representing $\{01,11\}$.

The traversal algorithm executes 3 iterations and terminate at 4th iteration. We list all intermediate 
results below:
\begin{itemize}
\item Iteration 1: $from^0 = \langle S+1+\alpha, x^2+x\rangle, to^1 = \langle T+1, x^2+x\rangle,
 reached = \langle T^2+\alpha T+(1+\alpha), x^2+x\rangle$
\item Iteration 2: $from^1 = \langle S+1, x^2+x\rangle, to^2= \langle T+\alpha, x^2+x\rangle,
reached = \langle T^3+1, x^2+x\rangle$
\item Iteration 3: $from^2 = \langle S+\alpha, x^2+x\rangle, to^3 = \langle T+\alpha x, x^2+x
\rangle, reached = \langle T^3\cdot x+x, T^4+T, x^2+x\rangle$
\item Iteration 4: $from^3 = \langle S, x\rangle, to^4 = \langle T+1, x\rangle, new = \langle1\rangle$
\end{itemize}
The final reachable states are represented by a multivariate polynomial ideal $\langle T^3\cdot x+x, T^4+T, x^2+x\rangle$,
which denotes $\{00,01,10,11\}$.
\end{Example}

@@

Figure \ref{fig:ZDD} shows a ZDD representing 
Equation \ref{eqn:chainOR} with size $2\times4 -1 = 7$, which is linear in the number of variables.
The reduction process using ZDDs can be executed as in \cite{PolyBoRi}. The graphical illustration 
of the remainder in ZDDs is also shown in Figure \ref{fig:ZDD}.

### 2016-12-31 dissertation/submission/unsat.tex
dissertation milestone: submitting to thesis office

\vspace{0.1in}
We compute its GB using Buchberger's algorithm with lexicographic term ordering $a>b>x>y>z$.
Since this problem is UNSAT, we will stop when ``1" is added to GB.
\begin{enumerate}[{1)}]
\item First we compute $Spoly(f_1,f_2)\xrightarrow{F}_{+} r_1$, remainder $r_1$ equals to $a$;
\item Update $F=F\cup r_1$;
\item Next we compute $Spoly(f_1,f_3)\xrightarrow{F}_{+} r_2$, remainder $r_2$ equals to $b$;
\item Update $F=F\cup r_2$;
\item We can use a directed acyclic graph (DAG) to represent the process to get $r_1,r_2$, as Figure \ref{fig:UNSAT}(a) shows;
\item Then we compute $Spoly(f_1,f_4) = s_3= a+b+1$, obviously $a+b+1$ can be reduced (multivariate divided) by
$r_1$ , the intermediate remainder $r_3 = b+1$. It can be immediately divided by $r_2$, and the remainder is ``1", we
terminate the Buchberger's algorithm;
\item We draw a DAG depicting the process through which we obtain remainder ``1" as shown in Figure \ref{fig:UNSAT}(b). 
From leaf ``1" we backtrace the graph to roots $f_1,f_2,f_3,f_4$. They constitute an UNSAT core for this problem
as these polynomials are the ``causes" of unsatisfiability of original set of polynomials.
\end{enumerate}
\end{Example}
\begin{figure}[h]
\centering{
\includegraphics[width=\textwidth]{newfig/UNSAT.eps}
\caption{DAG representing $Spoly$ computations and multivariate divisions.}}
\label{fig:UNSAT}
\end{figure}

@@

\begin{figure}[H]
\centering{
\includegraphics[width=\textwidth]{newfig/s27.eps}
\caption{Gate-level schematic of the example circuit.}}
\label{fig:s27}
\end{figure}

>>>>

\vspace{0.1in}
We compute its GB using Buchberger's algorithm with lexicographic term ordering $a>b>x>y>z$.
Since this problem is UNSAT, we will stop when ``1" is added to GB.
\begin{enumerate}[{1)}]
\item First we compute $Spoly(f_1,f_2)\xrightarrow{F}_{+} r_1$, remainder $r_1$ equals to $a$;
\item Update $F=F\cup r_1$;
\item Next we compute $Spoly(f_1,f_3)\xrightarrow{F}_{+} r_2$, remainder $r_2$ equals to $b$;
\item Update $F=F\cup r_2$;
\item We can use a directed acyclic graph (DAG) to represent the process to get $r_1,r_2$, as Figure \ref{fig:originUNSAT}(a) shows;
\item Then we compute $Spoly(f_1,f_4) = s_3= a+b+1$, obviously $a+b+1$ can be reduced (multivariate divided) by
$r_1$ , the intermediate remainder $r_3 = b+1$. It can be immediately divided by $r_2$, and the remainder is ``1", we
terminate the Buchberger's algorithm;
\item We draw a DAG depicting the process through which we obtain remainder ``1" as shown in Figure \ref{fig:originUNSAT}(b). 
From leaf ``1" we backtrace the graph to roots $f_1,f_2,f_3,f_4$. They constitute an UNSAT core for this problem
as these polynomials are the ``causes" of unsatisfiability of original set of polynomials.
\end{enumerate}
\end{Example}

\begin{figure}[h]
\centering{
\includegraphics[width=\textwidth]{newfig/UNSAT.eps}}
\caption{DAG representing $Spoly$ computations and multivariate divisions.}
\label{fig:originUNSAT}
\end{figure}

@@

\begin{figure}[H]
\centering{
\includegraphics[width=\textwidth]{newfig/s27.eps}}
\caption{Gate-level schematic of the example circuit.}
\label{fig:s27}
\end{figure}
